The role of aneuploidy in the evolution of cancer drug resistance

### 2024-02-15 EvolutionaryRescue1.tex
minor edits

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or ``barely growing'', that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment (we assume that aneuploidy significantly increases the death rate of wildtype cancer cells in an environment which is not affected by chemotherapeutic drugs). % TODO unclear how s affects \lambda_w; I do not think that s has any effect on \lambda_w as s is the fitness of the aneuploid in the drug free environment
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the fraction of cancer cells with favorable aneuploidy as $f=2/22\times0.15\approx 0.0136$ (i.e. $1.36\%$ of pre-treatment cancer cells have the beneficial aneuploidy). This gives us a fitness cost of $s= 7.3\times10^{-3}$ per day.
We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources. This is a good approximation as we assume that the drugs cause the wildtype population to decline rapidly and, as a result, density-dependent effects can be neglected.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate division and death rates that depend on the number of cells in the population, we assume the rates of division and death to be
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive wildtype cell without the involvement of aneuploidy, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wildtype cells are decreasing in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.  % TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one. so what are we talking about? try to rephrase. ok
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).  % TODO not sure what the last sentence "but ..." means and how do you see it ok
% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second case, $\frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. % NOTE this is the kind of conclusions we want to drive home
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\% 
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing the above approximation to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear ok
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

>>>>

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment (we assume that aneuploidy significantly increases the death rate of wildtype cancer cells in an environment which is not affected by chemotherapeutic drugs). % TODO unclear how s affects \lambda_w; I do not think that s has any effect on \lambda_w as s is the fitness of the aneuploid in the drug free environment
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the fraction of cancer cells with favorable aneuploidy as $f=2/22\times0.15\approx 0.0136$ (i.e. $1.36\%$ of pre-treatment cancer cells have the beneficial aneuploidy). This gives us a fitness cost of $s= 7.3\times10^{-3}$ per day.
\daniel{THIS DERIVATION DOESN'T MAKE SENSE. FROM THE LUKOW EXPERIMENT WE CAN DIRECTLY ESTIMATE THE FITNESS COST AS $s \approx \ln(0.5/0.15)/24 \approx .05$ PER DAY. DOING IT MORE CAREFULLY (LOGISTIC RATHER THAN EXPONENTIAL) GIVES ABOUT THE SAME ANSWER.}
We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analytical calculations, we assume that lineages produced by cells from the initial population divide and die independently of each other.
Of course, the cells will actually compete for resources, but we expect that this can be neglected because the drug will
cause the cell density to rapidly drop far below the carrying capacity at which these interactions are important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive wildtype cell without the involvement of aneuploidy, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w}  \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wildtype cells are decreasing in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.  % TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one. so what are we talking about? try to rephrase. ok
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).  % TODO not sure what the last sentence "but ..." means and how do you see it ok
% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second case, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. % NOTE this is the kind of conclusions we want to drive home
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could be evolutionarily rescued even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing the above approximation to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear ok
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
